---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Library

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(lattice)
```


## Loading and preprocessing the data

Imported the data file:

```{r}
unzip("activity.zip")
activity <- read.csv("activity.csv")
```

Tidy the data in preparation for analysis

```{r}
activity$date <- as.Date(activity$date, "%Y-%m-%d")
```


## What is mean total number of steps taken per day?

Calculating the total number of steps per day:

```{r}
total.steps <- tapply(activity$steps, as.factor(activity$date), sum, na.rm = TRUE)

#Total number of steps taken per day
total.steps
```

Plotting the average total steps per day as a histogram:

```{r}
hist(total.steps, main = "Total number of steps taken per day", xlab = "Total steps")
```

Calculating the mean and median of the total number of steps taken per day:

```{r}
meansteps <- mean(total.steps, na.rm = TRUE)
mediansteps <- median(total.steps, na.rm = TRUE)
```

The mean of the total number of steps taken per day is `r meansteps`, and the median is `r mediansteps` 

## What is the average daily activity pattern?

Making a time series plot of the 5-minute interval and the average steps taken, averaged across all days:

```{r}
# Calculating the average steps per interval
steps.interval <- activity %>%
  group_by(interval) %>%
  summarise(across(steps, mean, na.rm = TRUE))

# Plotting results as a time series
plot(steps.interval$interval, steps.interval$steps, type = "l", main = "Time series plot of steps per interval", xlab = "Interval", ylab = "Steps")

# The 5 minute interval with the maximum number of steps
steps.interval[which.max(steps.interval$steps),]
maxsteps <- max(steps.interval$steps)
```

The 5 minute interval which contains the most number of steps on average is the interval from 08:35-08:40 in the morning. An average of `r maxsteps` is taken in this interval.  

## Imputing missing values 

Calculating the total number of missing values in the dataset:

```{r}
missing <- sum(is.na(activity))
```

There are a total of `r missing` values in the dataset.

Filling in the missing values with the mean steps per interval:

```{r}
activity %>%
  group_by(interval) %>%
  mutate(steps = replace_na(steps, mean(steps, na.rm = TRUE)))

#Creating a new dataset with missing data filled in

activity.imputed <- activity %>%
  group_by(interval) %>%
  mutate(steps = replace_na(steps, mean(steps, na.rm = TRUE)))
```

Creating a histogram of the total number of steps, and calculating the mean and median (after imputation):

```{r}
# Histogram of total steps

total.steps.imputed <- tapply(activity.imputed$steps, as.factor(activity.imputed$date), sum)

hist(total.steps.imputed)

# Mean and median

meansteps.imputed <- mean(total.steps.imputed)
mediansteps.imputed <- median(total.steps.imputed)
```

After imputation, the total steps per day becomes more normally distributed. The mean and median values both increase.

## Are there differences in activity patterns between weekdays and weekends?

Creating a new factor variable indicating which days are weekdays:

```{r}
activity.imputed <- activity.imputed %>%
  ungroup() %>%
  mutate(day.type = weekdays(activity.imputed$date)) %>%
  mutate(day.type = recode(day.type, "Monday" = "weekday",
                           "Tuesday" = "weekday",
                           "Wednesday" = "weekday",
                           "Thursday" = "weekday",
                           "Friday" = "weekday",
                           "Saturday" = "weekend",
                           "Sunday" = "weekend"))

```

Plotting the activity pattern:

```{r}
# Calculating the average steps per interval
steps.interval.imputed <- activity.imputed %>%
  group_by(interval, day.type) %>%
  summarise(across(steps, mean, na.rm = TRUE))

# Making the plot
xyplot(steps ~ interval | day.type, data = steps.interval.imputed, type = "l", layout = c(1, 2))
```


